---
output: 
  html_document:
    css: styles.css
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

<h2> Comparing PRISM and DayMet with weather station data. </h2>

<ul>
<li>Weather station data were downloaded from the Applied Climate Information System API (www.rcc-acis.org) for the McFarland Hill WS in Acadia NP. </li>
<li>Prism data are 4km2 rasters that were downloaded using the prism R package then values for each plot were extracted. </li>
<li>DayMet data were batch downloaded for each point using the daymetr package</li>
<li>Standardized Precipitation-Evapotranspiration Index (SPEI) was calculated using DayMet data and Hargraeves PET using the SPEI R package.</li>
<li>Global 1 degree SPEI data were downloaded from the SPEI database (spei.csic.es) and extracted for each point in R.</li>
<li>Code: https://github.com/KateMMiller/tree_ring_gamms/blob/main/01B_compile_climate_atmdep_data.R</li>
</ul>

Based on these comparisons, I decided to use DayMet data for analyses. 
<ul><li>Temperature ranges across all plots within a year were consistently lower in PRISM than DayMet, but I don't think climate varies quite as much across the park PRISM assumes.</li>
<li>Correlations between DayMet and weather station temperature data were much stronger than PRISM.</li>
</ul>
```{r import, include = FALSE}
library(tidyverse)
library(GGally)
clim_dat <- read.csv("./data/temp/ACAD_climate_data.csv") |> filter(year >=1980)

```

```{r results = 'hide'}
#---- Compare weather station, prism and daymet data for near and far plots ----
tmax7_long <- clim_dat |> filter(Plot_Name %in% c("ACAD-039", "ACAD-064", "ACAD-036")) |> 
  select(Plot_Name, year, ws_tmaxc_7, dm_tmax_mean_7, pr_tmax_07) |> 
  pivot_longer(ws_tmaxc_7:pr_tmax_07, names_to = "station", values_to = "tmax_7")

```

<h3>Plot-Level Comparisons</h3>
<details open><summary class='drop'>Average Maximum Temperature in July (C)</summary>
```{r message = FALSE, warning = FALSE, fig.height = 6}
tmax_july_p39 <- 
ggplot(tmax7_long |> filter(Plot_Name == "ACAD-039"), 
       aes(x = year, y = tmax_7, group = station, color = station, shape = station, fill = station)) +
  geom_point() + geom_smooth(se = F) + forestNETN::theme_FHM() + theme(legend.position = 'none') + 
  labs(y = "July Max Temp. C", x = NULL, subtitle = "ACAD-039: closest to WS") +
  scale_fill_manual(values = c("dm_tmax_mean_7" = "orange", 
                               "pr_tmax_07" = "dodgerblue", 
                               "ws_tmaxc_7" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_color_manual(values = c("dm_tmax_mean_7" = "orange", 
                                "pr_tmax_07" = "dodgerblue", 
                                "ws_tmaxc_7" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_shape_manual(values = c("dm_tmax_mean_7" = 23, 
                               "pr_tmax_07" = 21, 
                               "ws_tmaxc_7" = 24), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_size_manual(values = c("dm_tmax_mean_7" = 3.5, 
                               "pr_tmax_07" = 3.5, 
                               "ws_tmaxc_7" = 3), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) 

tmax_july_p36 <- 
ggplot(tmax7_long |> filter(Plot_Name == "ACAD-036"), 
       aes(x = year, y = tmax_7, group = station, color = station, shape = station, fill = station)) +
  geom_point() + geom_smooth(se = F) + forestNETN::theme_FHM() + theme(legend.position = 'none') + 
  labs(y = "July Max Temp. C",  x = NULL, subtitle = "ACAD-036: SCH farthest east of WS") +
  scale_fill_manual(values = c("dm_tmax_mean_7" = "orange", 
                               "pr_tmax_07" = "dodgerblue", 
                               "ws_tmaxc_7" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_color_manual(values = c("dm_tmax_mean_7" = "orange", 
                                "pr_tmax_07" = "dodgerblue", 
                                "ws_tmaxc_7" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_shape_manual(values = c("dm_tmax_mean_7" = 23, 
                               "pr_tmax_07" = 21, 
                               "ws_tmaxc_7" = 24), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_size_manual(values = c("dm_tmax_mean_7" = 3.5, 
                               "pr_tmax_07" = 3.5, 
                               "ws_tmaxc_7" = 3), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) 
  
tmax_july_p64 <- 
ggplot(tmax7_long |> filter(Plot_Name == "ACAD-064"), 
       aes(x = year, y = tmax_7, group = station, color = station, shape = station, fill = station)) +
  geom_point() + geom_smooth(se = F) + forestNETN::theme_FHM() + theme(legend.position = 'bottom') +
  labs(y = "July Max Temp. C",  x = NULL, subtitle = "ACAD-064: IAH farthest west of WS") +
  scale_fill_manual(values = c("dm_tmax_mean_7" = "orange", 
                               "pr_tmax_07" = "dodgerblue", 
                               "ws_tmaxc_7" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_color_manual(values = c("dm_tmax_mean_7" = "orange", 
                                "pr_tmax_07" = "dodgerblue", 
                                "ws_tmaxc_7" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_shape_manual(values = c("dm_tmax_mean_7" = 23, 
                               "pr_tmax_07" = 21, 
                               "ws_tmaxc_7" = 24), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_size_manual(values = c("dm_tmax_mean_7" = 3.5, 
                               "pr_tmax_07" = 3.5, 
                               "ws_tmaxc_7" = 3), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) 

cowplot::plot_grid(tmax_july_p39, tmax_july_p36, tmax_july_p64, ncol = 1, 
                   rel_heights = c(0.3, 0.3, 0.4))

```
</details>
<details open><summary class='drop'>Average Minimum Temperature in January (C)</summary>
```{r message = FALSE, warning = FALSE, fig.height = 6}
tmin1_long <- clim_dat |> filter(Plot_Name %in% c("ACAD-039", "ACAD-064", "ACAD-036")) |> 
  select(Plot_Name, year, ws_tminc_1, dm_tmin_mean_1, pr_tmin_01) |> 
  pivot_longer(ws_tminc_1:pr_tmin_01, names_to = "station", values_to = "tmin_1")

tmin_jan_p39 <- 
ggplot(tmin1_long |> filter(Plot_Name == "ACAD-039"), 
       aes(x = year, y = tmin_1, group = station, color = station, shape = station, fill = station)) +
  geom_point() + geom_smooth(se = F) + forestNETN::theme_FHM() + theme(legend.position = 'none') + 
  labs(y = "Jan. Min. Temp. C",  x = NULL, subtitle = "ACAD-039: closest to WS") +
  scale_fill_manual(values = c("dm_tmin_mean_1" = "orange", 
                               "pr_tmin_01" = "dodgerblue", 
                               "ws_tminc_1" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_color_manual(values = c("dm_tmin_mean_1" = "orange", 
                                "pr_tmin_01" = "dodgerblue", 
                                "ws_tminc_1" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_shape_manual(values = c("dm_tmin_mean_1" = 23, 
                               "pr_tmin_01" = 21, 
                               "ws_tminc_1" = 24), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_size_manual(values = c("dm_tmin_mean_1" = 3.5, 
                               "pr_tmin_01" = 3.5, 
                               "ws_tminc_1" = 3), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) 

tmin_jan_p36 <- 
  ggplot(tmin1_long |> filter(Plot_Name == "ACAD-036"), 
       aes(x = year, y = tmin_1, group = station, color = station, shape = station, fill = station)) +
  geom_point() + geom_smooth(se = F) + forestNETN::theme_FHM() + theme(legend.position = 'none') + 
  labs(y = "Jan. Min. Temp. C",  x = NULL, subtitle = "ACAD-036: SCH farthest east from WS") +
  scale_fill_manual(values = c("dm_tmin_mean_1" = "orange", 
                               "pr_tmin_01" = "dodgerblue", 
                               "ws_tminc_1" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_color_manual(values = c("dm_tmin_mean_1" = "orange", 
                                "pr_tmin_01" = "dodgerblue", 
                                "ws_tminc_1" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_shape_manual(values = c("dm_tmin_mean_1" = 23, 
                               "pr_tmin_01" = 21, 
                               "ws_tminc_1" = 24), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_size_manual(values = c("dm_tmin_mean_1" = 3.5, 
                               "pr_tmin_01" = 3.5, 
                               "ws_tminc_1" = 3), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) 


tmin_jan_p64 <- 
  ggplot(tmin1_long |> filter(Plot_Name == "ACAD-064"), 
       aes(x = year, y = tmin_1, group = station, color = station, shape = station, fill = station)) +
  geom_point() + geom_smooth(se = F) + forestNETN::theme_FHM() + theme(legend.position = 'bottom') + 
  labs(y = "Jan. Min. Temp. C",  x = NULL, subtitle = "ACAD-064: IAH farthest west from WS") +
  scale_fill_manual(values = c("dm_tmin_mean_1" = "orange", 
                               "pr_tmin_01" = "dodgerblue", 
                               "ws_tminc_1" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_color_manual(values = c("dm_tmin_mean_1" = "orange", 
                                "pr_tmin_01" = "dodgerblue", 
                                "ws_tminc_1" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_shape_manual(values = c("dm_tmin_mean_1" = 23, 
                               "pr_tmin_01" = 21, 
                               "ws_tminc_1" = 24), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_size_manual(values = c("dm_tmin_mean_1" = 3.5, 
                               "pr_tmin_01" = 3.5, 
                               "ws_tminc_1" = 3), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) 

cowplot::plot_grid(tmin_jan_p39, tmin_jan_p36, tmin_jan_p64, ncol = 1, 
                   rel_heights = c(0.3, 0.3, 0.4))

```
</details>
<details open><summary class='drop'>Total July Precip. (mm)</summary>
```{r message = FALSE, warning = FALSE, fig.height = 6}
ppt7_long <- clim_dat |> filter(Plot_Name %in% c("ACAD-039", "ACAD-064", "ACAD-036")) |> 
  select(Plot_Name, year, ws_pcpmm_7, dm_ppt_sum_7, pr_ppt_07) |> 
  pivot_longer(ws_pcpmm_7:pr_ppt_07, names_to = "station", values_to = "ppt_7")

ppt_jul_p39 <- 
ggplot(ppt7_long |> filter(Plot_Name == "ACAD-039"), 
       aes(x = year, y = ppt_7, group = station, color = station, shape = station, fill = station)) +
  geom_point() + geom_smooth(se = F) + forestNETN::theme_FHM() + theme(legend.position = 'none') + 
  labs(y = "July Precip.",  x = NULL, subtitle = "ACAD-039: closest to WS") +
  scale_fill_manual(values = c("dm_ppt_sum_7" = "orange", 
                               "pr_ppt_07" = "dodgerblue", 
                               "ws_pcpmm_7" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_color_manual(values = c("dm_ppt_sum_7" = "orange", 
                                "pr_ppt_07" = "dodgerblue", 
                                "ws_pcpmm_7" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_shape_manual(values = c("dm_ppt_sum_7" = 23, 
                               "pr_ppt_07" = 21, 
                               "ws_pcpmm_7" = 24), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_size_manual(values = c("dm_ppt_sum_7" = 3.5, 
                               "pr_ppt_07" = 3.5, 
                               "ws_pcpmm_7" = 3), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) 

ppt_jul_p36 <- 
  ggplot(ppt7_long |> filter(Plot_Name == "ACAD-036"), 
       aes(x = year, y = ppt_7, group = station, color = station, shape = station, fill = station)) +
  geom_point() + geom_smooth(se = F) + forestNETN::theme_FHM() + theme(legend.position = 'none') + 
  labs(y = "July Precip.",  x = NULL, subtitle = "ACAD-036: SCH farthest east from WS") +
  scale_fill_manual(values = c("dm_ppt_sum_7" = "orange", 
                               "pr_ppt_07" = "dodgerblue", 
                               "ws_pcpmm_7" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_color_manual(values = c("dm_ppt_sum_7" = "orange", 
                                "pr_ppt_07" = "dodgerblue", 
                                "ws_pcpmm_7" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_shape_manual(values = c("dm_ppt_sum_7" = 23, 
                               "pr_ppt_07" = 21, 
                               "ws_pcpmm_7" = 24), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_size_manual(values = c("dm_ppt_sum_7" = 3.5, 
                               "pr_ppt_07" = 3.5, 
                               "ws_pcpmm_7" = 3), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) 


ppt_jul_p64 <- 
  ggplot(ppt7_long |> filter(Plot_Name == "ACAD-064"), 
       aes(x = year, y = ppt_7, group = station, color = station, shape = station, fill = station)) +
  geom_point() + geom_smooth(se = F) + forestNETN::theme_FHM() + theme(legend.position = 'bottom') + 
  labs(y = "July Precip.",  x = NULL, subtitle = "ACAD-064: IAH farthest west from WS") +
  scale_fill_manual(values = c("dm_ppt_sum_7" = "orange", 
                               "pr_ppt_07" = "dodgerblue", 
                               "ws_pcpmm_7" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_color_manual(values = c("dm_ppt_sum_7" = "orange", 
                                "pr_ppt_07" = "dodgerblue", 
                                "ws_pcpmm_7" = "#949494"), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_shape_manual(values = c("dm_ppt_sum_7" = 23, 
                               "pr_ppt_07" = 21, 
                               "ws_pcpmm_7" = 24), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) +
  scale_size_manual(values = c("dm_ppt_sum_7" = 3.5, 
                               "pr_ppt_07" = 3.5, 
                               "ws_pcpmm_7" = 3), 
                    name = "Data Type",
                    labels = c("DayMet","Prism", "Weather Station")) 

cowplot::plot_grid(ppt_jul_p39, ppt_jul_p36, ppt_jul_p64, ncol = 1, 
                   rel_heights = c(0.3, 0.3, 0.4))
```
</details>

<h3>Overall Comparisons</h3>
<details open><summary class='drop'>Average Maximum Temperature in July (C)</summary>
```{r, message = F, warning = F, fig.height = 4}
ggplot(clim_dat, aes(x = ws_tmaxc_7)) + forestNETN::theme_FHM() +
  geom_point(aes(y = dm_tmax_mean_7, fill = "DayMet", shape = "DayMet"), size = 2) +
  geom_point(aes(y = pr_tmax_07, fill = "Prism", shape = "Prism"), size = 2) +
  scale_fill_manual(values = c("orange", "dodgerblue"),
                    name = "Data Type",
                    labels = c("DayMet", "Prism")) +
  scale_shape_manual(values = c(23, 21),
                     name = "Data Type",
                     labels = c("DayMet", "Prism")) +
  labs(x = "Weather Station Data", y = "Raster Data", subtitle = "Max. July Temp.") +
  geom_abline(slope = 1, intercept = 0) + theme(aspect.ratio = 1, legend.title = element_blank()) 


```
</details>

<details open><summary class='drop'>Avg. Max July Temp. Correlations</summary>
```{r}
ggpairs(clim_dat |> select(WS = ws_tmaxc_7, DayMet = dm_tmax_mean_7, Prism = pr_tmax_07)) + 
  forestNETN::theme_FHM() +
  labs(title = "July Max Temp.")
```
</details>

<details open><summary class='drop'>Average Minimum Temperature in January (C)</summary>
```{r, fig.height = 4}
ggplot(clim_dat, aes(x = ws_tminc_1)) + forestNETN::theme_FHM() +
  geom_point(aes(y = dm_tmin_mean_1, fill = "DayMet", shape = "DayMet"), size = 2) +
  geom_point(aes(y = pr_tmin_01, fill = "Prism", shape = "Prism"), size = 2.5) +
  scale_fill_manual(values = c("orange", "dodgerblue"),
                    name = "Data Type",
                    labels = c("DayMet", "Prism")) +
  scale_shape_manual(values = c(23, 21),
                     name = "Data Type",
                     labels = c("DayMet", "Prism")) +
  labs(x = "Weather Station Data", y = "Raster Data", subtitle = "Min. January Temp.") +
  geom_abline(slope = 1, intercept = 0) + theme(aspect.ratio = 1, legend.title = element_blank()) 


```
</details>

<details open><summary class='drop'>Avg. Min January Temp. Correlations</summary>
```{r}
ggpairs(clim_dat |> select(WS = ws_tminc_1, DayMet = dm_tmin_mean_1, Prism = pr_tmin_01)) + forestNETN::theme_FHM() +
  labs(title = "Jan. Min Temp.")
```
</details>


<details open><summary class='drop'>Total July Precip. (mm)</summary>
```{r, fig.height = 4}
ggplot(clim_dat, aes(x = ws_pcpmm_7)) + forestNETN::theme_FHM() +
  geom_point(aes(y = dm_ppt_sum_7, fill = "DayMet", shape = "DayMet"), size = 2) +
  geom_point(aes(y = pr_ppt_07, fill = "Prism", shape = "Prism"), size = 2.5) +
  scale_fill_manual(values = c("orange", "dodgerblue"),
                    name = "Data Type",
                    labels = c("DayMet", "Prism")) +
  scale_shape_manual(values = c(23, 21),
                     name = "Data Type",
                     labels = c("DayMet", "Prism")) +
  labs(x = "Weather Station Data", y = "Raster Data", subtitle = "July Precip. (mm)") +
  geom_abline(slope = 1, intercept = 0) + theme(aspect.ratio = 1, legend.title = element_blank()) 


```
</details>

<details open><summary class='drop'>July Precip. Correlations</summary>
```{r}
ggpairs(clim_dat |> select(WS = ws_pcpmm_7, DayMet = dm_ppt_sum_7, Prism = pr_ppt_07)) + forestNETN::theme_FHM() +
  labs(title = "July Precip")
```
</details>


<h3> Standardized Precipitation-Evapotranspiration Index (SPEI)</h3>
<details open><summary class='drop'>August 1-month SPEI</summary>
```{r, fig.height = 4}
ggplot(clim_dat, aes(x = year)) + forestNETN::theme_FHM() +
  geom_point(aes(y = dm_SPEI01_8, color = "DayMet", shape = "DayMet"), size = 2) +
  geom_smooth(aes(x = year, y = dm_SPEI01_8), color = 'dodgerblue3', se = F) +
  geom_point(aes(y = nc_SPEI01_8, color = "Global", shape = "Global"), size = 2) +
  geom_smooth(aes(x = year, y = nc_SPEI01_8), color = "#D07712", se = F) +
  # scale_fill_manual(values = c("#F7D196", "#A3CDFC"),
  #                   name = "Data Type",
  #                   labels = c("DayMet", "Global")) +
  scale_color_manual(values = c("orange", "dodgerblue"),
                    name = "Data Type",
                    labels = c("DayMet", "Global")) +
  scale_shape_manual(values = c(18, 19),
                     name = "Data Type",
                     labels = c("DayMet", "Global")) +
  labs(x = "Year", y = "August SPEI-1 month") +
  geom_abline(slope = 1, intercept = 0) + theme(aspect.ratio = 1, legend.title = element_blank()) 
```
</details>

<details open><summary class='drop'>August 3-month SPEI</summary>
```{r, fig.height = 4}
ggplot(clim_dat, aes(x = year)) + forestNETN::theme_FHM() +
  geom_point(aes(y = dm_SPEI03_8, color = "DayMet", shape = "DayMet"), size = 2) +
  geom_smooth(aes(x = year, y = dm_SPEI03_8), color = 'dodgerblue3', se = F) +
  geom_point(aes(y = nc_SPEI03_8, color = "Global", shape = "Global"), size = 2) +
  geom_smooth(aes(x = year, y = nc_SPEI03_8), color = "#D07712", se = F) +
  # scale_fill_manual(values = c("#F7D196", "#A3CDFC"),
  #                   name = "Data Type",
  #                   labels = c("DayMet", "Global")) +
  scale_color_manual(values = c("orange", "dodgerblue"),
                    name = "Data Type",
                    labels = c("DayMet", "Global")) +
  scale_shape_manual(values = c(18, 19),
                     name = "Data Type",
                     labels = c("DayMet", "Global")) +
  labs(x = "Year", y = "August SPEI-3 month") +
  geom_abline(slope = 1, intercept = 0) + theme(aspect.ratio = 1, legend.title = element_blank()) 

```
</details>
