---
output: 
  html_document:
    css: styles.css
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

#### Comparing PRISM and DayMet with weather station data. 

<ul>
<li>Weather station data were downloaded from the Applied Climate Information System API (www.rcc-acis.org) for the McFarland Hill WS in Acadia NP. </li>
<li>Prism data are 4km2 rasters that were downloaded using the prism R package then values for each plot were extracted. </li>
<li>DayMet data were batch downloaded for each point using the daymetr package</li>
<li>Standardized Precipitation-Evapotranspiration Index (SPEI) was calculated using DayMet data and Hargraeves PET using the SPEI R package.</li>
<li>Global 1 degree SPEI data were downloaded from the SPEI database (spei.csic.es) and extracted for each point in R.</li>
<li>Code: https://github.com/KateMMiller/tree_ring_gamms/blob/main/01B_compile_climate_atmdep_data.R</li>
</ul>

```{r import, include = FALSE}
library(tidyverse)
clim_dat <- read.csv("./data/temp/ACAD_climate_data.csv")

```

```{r results = 'hide'}
#---- Compare weather station, prism and daymet data for near and far plots ----
ws_comp_39 <- clim_dat |> #filter(between(year, 1985, 2014)) |> 
  filter(Plot_Name == "ACAD-039") # closest to WS
ws_comp_64 <- clim_dat |> #filter(between(year, 1985, 2014)) |> 
  filter(Plot_Name == "ACAD-064") # IAH- farthest
ws_comp_36 <- clim_dat |> #filter(between(year, 1985, 2014)) |> 
  filter(Plot_Name == "ACAD-036") # SCH- farthest

# Function for comparing weather station, daymet and prism data for 1 plot of data
ws_dm_pr_plot <- function(dat, ws, dm, pr, ylab, title, legend = "none"){
  ggplot(dat, aes(x = year)) + forestNETN::theme_FHM() + 
  geom_smooth(aes(x = year, y = {{ws}}, color = 'Weather Station'), se = F) +
  geom_point(aes(y = {{ws}}, fill = "Weather station", shape = "Weather station", size = "Weather station")) +
  geom_smooth(aes(x = year, y = {{dm}}, color = 'DayMet'), se = F) +
  geom_point(aes(y = {{dm}}, fill = "DayMet", shape = "DayMet", size = "DayMet")) +
  geom_smooth(aes(x = year, y = {{pr}}, color = "Prism"), se = F) +
  geom_point(aes(y = {{pr}}, fill = "Prism", shape = "Prism", size = "Prism")) +
  
  labs(y = ylab, subtitle = title) +
  scale_fill_manual(values = c("#949494", "orange", "dodgerblue"), 
                    name = "Data Type",
                    labels = c("Weather Station", "DayMet", "Prism")) +
  scale_color_manual(values = c("#949494", "orange", "dodgerblue"), 
                    name = "Data Type",
                    labels = c("Weather Station", "DayMet", "Prism")) +
  scale_shape_manual(values = c(24, 23, 21), 
                     name = "Data Type",
                     labels = c("Weather Station", "DayMet", "Prism")) +
  scale_size_manual(values = c(3, 3.5, 3.5), 
                    name = "Data Type",
                    labels = c("Weather Station", "DayMet", "Prism")) +
  theme(legend.position = legend, legend.title = element_blank())
}

```

<details open><summary><h4>Plot-Level Comparisons</h4></summary>
<details open><summary><h5>Average Maximum Temperature in July (C)</h5></summary>
```{r message = FALSE, warning = FALSE, height = 6}
tmax_july_p39 <- 
ws_dm_pr_plot(ws_comp_39, ws = ws_tmaxc_7, dm = dm_tmax_mean_7, pr = pr_tmax_07,
              ylab = "Max Temp. (C)", title = "ACAD-039: closest to WS")

tmax_july_p36 <- 
ws_dm_pr_plot(ws_comp_36, ws = ws_tmaxc_7, dm = dm_tmax_mean_7, pr = pr_tmax_07,
              ylab = "Max Temp. (C)", title = "ACAD-036: SCH far from WS")

tmax_july_p64 <- 
ws_dm_pr_plot(ws_comp_64, ws = ws_tmaxc_7, dm = dm_tmax_mean_7, pr = pr_tmax_07,
              ylab = "Max Temp. (C)", title = "ACAD-064: IAH far from WS", 
              legend = 'bottom')

cowplot::plot_grid(tmax_july_p39, tmax_july_p36, tmax_july_p64, ncol = 1, 
                   rel_heights = c(0.3, 0.3, 0.4))

```
</details>
<details open><summary><h5>Average Minimum Temperature in January (C)</h5></summary>
```{r message = FALSE, warning = FALSE, height = 6}
tmin_jan_p39 <- 
ws_dm_pr_plot(ws_comp_39, ws = ws_tminc_1, dm = dm_tmin_mean_1, pr = pr_tmin_01,
              ylab = "Min. Temp. (C)", title = "ACAD-039: closest to WS")

tmin_jan_p36 <- 
ws_dm_pr_plot(ws_comp_36, ws = ws_tminc_1, dm = dm_tmin_mean_1, pr = pr_tmin_01,
              ylab = "Min. Temp. (C)", title = "ACAD-036: SCH far from WS")

tmin_jan_p64 <- 
ws_dm_pr_plot(ws_comp_64, ws = ws_tminc_1, dm = dm_tmin_mean_1, pr = pr_tmin_01,
              ylab = "Min. Temp. (C)", title = "ACAD-064: IAH far from WS", 
              legend = 'bottom')

cowplot::plot_grid(tmin_jan_p39, tmin_jan_p36, tmin_jan_p64, ncol = 1, 
                   rel_heights = c(0.3, 0.3, 0.4))

```
</details>
<details open><summary><h5>Total July Precip. (mm)</h5></summary>
```{r message = FALSE, warning = FALSE, height = 6}
ppt_july_p39 <- 
ws_dm_pr_plot(ws_comp_39, ws = ws_pcpmm_7, dm = dm_ppt_sum_7, pr = pr_ppt_07,
              ylab = "July Precip. (mm)", title = "ACAD-039: closest to WS")

ppt_july_p36 <- 
ws_dm_pr_plot(ws_comp_36, ws = ws_pcpmm_7, dm = dm_ppt_sum_7, pr = pr_ppt_07,
              ylab = "July Precip. (mm)", title = "ACAD-036: SCH far from WS")

ppt_july_p64 <- 
ws_dm_pr_plot(ws_comp_64, ws = ws_pcpmm_7, dm = dm_ppt_sum_7, pr = pr_ppt_07,
              ylab = "July Precip. (mm)", title = "ACAD-064: IAH far from WS", legend = 'bottom')

cowplot::plot_grid(ppt_july_p39, ppt_july_p36, ppt_july_p64, ncol = 1,
                   rel_heights = c(0.3, 0.3, 0.4))
```
</details></details>

<details open><summary><h4>Overall Comparisons</h4></summary>
<details open><summary><h5>Average Maximum Temperature in July (C)</h5></summary>
```{r, message = F, warning = F}
ggplot(clim_dat, aes(x = ws_tmaxc_7)) + forestNETN::theme_FHM() +
  geom_point(aes(y = dm_tmax_mean_7, fill = "DayMet", shape = "DayMet"), size = 2) +
  geom_point(aes(y = pr_tmax_07, fill = "Prism", shape = "Prism"), size = 2) +
  scale_fill_manual(values = c("orange", "dodgerblue"),
                    name = "Data Type",
                    labels = c("DayMet", "Prism")) +
  scale_shape_manual(values = c(23, 21),
                     name = "Data Type",
                     labels = c("DayMet", "Prism")) +
  labs(x = "Weather Station Data", y = "Raster Data", subtitle = "Max. July Temp.") +
  geom_abline(slope = 1) + theme(aspect.ratio = 1, legend.title = element_blank()) 


```
</details>

<details open><summary><h5>Average Minimum Temperature in January (C)</h5></summary>
```{r}
ggplot(clim_dat, aes(x = ws_tminc_1)) + forestNETN::theme_FHM() +
  geom_point(aes(y = dm_tmin_mean_1, fill = "DayMet", shape = "DayMet"), size = 2) +
  geom_point(aes(y = pr_tmin_01, fill = "Prism", shape = "Prism"), size = 2.5) +
  scale_fill_manual(values = c("orange", "dodgerblue"),
                    name = "Data Type",
                    labels = c("DayMet", "Prism")) +
  scale_shape_manual(values = c(23, 21),
                     name = "Data Type",
                     labels = c("DayMet", "Prism")) +
  labs(x = "Weather Station Data", y = "Raster Data", subtitle = "Min. January Temp.") +
  geom_abline(slope = 1) + theme(aspect.ratio = 1, legend.title = element_blank()) 


```
</details>

<details open><summary><h5>Total July Precip. (mm)</h5></summary>
```{r}
ggplot(clim_dat, aes(x = ws_pcpmm_7)) + forestNETN::theme_FHM() +
  geom_point(aes(y = dm_ppt_sum_7, fill = "DayMet", shape = "DayMet"), size = 2) +
  geom_point(aes(y = pr_ppt_07, fill = "Prism", shape = "Prism"), size = 2.5) +
  scale_fill_manual(values = c("orange", "dodgerblue"),
                    name = "Data Type",
                    labels = c("DayMet", "Prism")) +
  scale_shape_manual(values = c(23, 21),
                     name = "Data Type",
                     labels = c("DayMet", "Prism")) +
  labs(x = "Weather Station Data", y = "Raster Data", subtitle = "July Precip. (mm)") +
  geom_abline(slope = 1) + theme(aspect.ratio = 1, legend.title = element_blank()) 


```
</details></details>

<details open><summary><h4> Standardized Precipitation-Evapotranspiration Index (SPEI)</h4></summary>
<details open><summary><h5>August 1-month SPEI</h5></summary>
```{r}
ggplot(clim_dat, aes(x = year)) + forestNETN::theme_FHM() +
  geom_point(aes(y = dm_SPEI01_8, color = "DayMet", shape = "DayMet"), size = 2) +
  geom_smooth(aes(x = year, y = dm_SPEI01_8), color = 'dodgerblue3', se = F) +
  geom_point(aes(y = nc_SPEI01_8, color = "Global", shape = "Global"), size = 2) +
  geom_smooth(aes(x = year, y = nc_SPEI01_8), color = "#D07712", se = F) +
  # scale_fill_manual(values = c("#F7D196", "#A3CDFC"),
  #                   name = "Data Type",
  #                   labels = c("DayMet", "Global")) +
  scale_color_manual(values = c("orange", "dodgerblue"),
                    name = "Data Type",
                    labels = c("DayMet", "Global")) +
  scale_shape_manual(values = c(18, 19),
                     name = "Data Type",
                     labels = c("DayMet", "Global")) +
  labs(x = "Year", y = "August SPEI-1 month") +
  geom_abline(slope = 1) + theme(aspect.ratio = 1, legend.title = element_blank()) 
```
</details>

<details open><summary><h5>August 3-month SPEI</h5></summary>
```{r}
ggplot(clim_dat, aes(x = year)) + forestNETN::theme_FHM() +
  geom_point(aes(y = dm_SPEI03_8, color = "DayMet", shape = "DayMet"), size = 2) +
  geom_smooth(aes(x = year, y = dm_SPEI03_8), color = 'dodgerblue3', se = F) +
  geom_point(aes(y = nc_SPEI03_8, color = "Global", shape = "Global"), size = 2) +
  geom_smooth(aes(x = year, y = nc_SPEI03_8), color = "#D07712", se = F) +
  # scale_fill_manual(values = c("#F7D196", "#A3CDFC"),
  #                   name = "Data Type",
  #                   labels = c("DayMet", "Global")) +
  scale_color_manual(values = c("orange", "dodgerblue"),
                    name = "Data Type",
                    labels = c("DayMet", "Global")) +
  scale_shape_manual(values = c(18, 19),
                     name = "Data Type",
                     labels = c("DayMet", "Global")) +
  labs(x = "Year", y = "August SPEI-3 month") +
  geom_abline(slope = 1) + theme(aspect.ratio = 1, legend.title = element_blank()) 

```
</details></details>
