---
title: "tree ring gam notes"
output: html_notebook
editor_options: 
  chunk_output_type: console
---
Referencing Circuzzi et al. 2021:
- "Tree cores with low interseries correlation among the sampled population are typically removed." Make sure I check for this.
- Calculated basal area increment, the annual increase in cross-sectional area of a tree to obtain a two-dimensional estimate of annual growth for each tree


Model covariates
X month
X year
X crown class (codom or inter)
X DBH 
X species
X elevation
X slope
X aspect
X Unit (MDI-E, MDI-W, Schoodic, IAH)
X Vegetation type (use Group_1 from shapefile)
X soil pH (of first sample)
X % base saturation (of first sample)
X % TC (of first sample)
X % TN (of first sample)
X forest floor soil depth (0-10) of first sample
X Fire 1947 (binary)
X SPEI: standardized precipitation-evapotranspiration index (3 month metric from PRISM)
X Deposition
X TMIN/TMAX
X Precip (whatever Paula used)
  ? growing season precip and max temperature
  ? number of drought days during growing season
  ? longest number of consecutive days in drought
Data cleaning tasks
X Take ringwidth data and reshape so each row is a core and each column is a year
X Pull out plot number from the ring-width data 

Notes: 
- Building out a model with full range of years to get a feel for the mechanics of the gamms. For the real analysis, will likely truncate around mid 1980s when atmospheric deposition data are available to test how well various models on deposition, climate, and site-level covariates predict growth.
- Replication of cores within plot is <5, and often closer to 1 or 2 of a species, which is not enough to estimate within plot variance for random effects. Instead, I'll randomly select one core from each site for the analysis, then use the unselected cores as part of the validation process.
- I initially tried modelling all species at the same time, but the models were too complex/cumbersome, so I'm sticking with species-specific models. This also allows us to test different models (or see how a set of models predicts growth of species differently).
- Ultimately, I'd like to move this into a Bayesian framework, as I find it easier to set up the different model levels (eg variables that are at the plot-level and unchanging across years, variables that have a value for every year but that don't vary by plot).
- Check PRISM data to get individual values for each site, rather than weather station data that's the same for entire park.


```{r import}
library(tidyverse)
library(mgcv)
library(gratia) # for easy plotting of results and diagnostics
library(mgcViz) # for more customized gam plotting
library(broom) # for tidy model output

cordat_pre <- read.csv("./data/Full_plot_core_climate_deposition_dataset.csv") 
table(cordat_pre$Crown_Class) # 118 records with Open Grown (1); 124 with Dominant (2)

cordat <- cordat_pre %>% filter(Crown_Class %in% c(3, 4)) 

head(cordat)
str(cordat[,1:30])
names(cordat)
# have to convert to factors, otherwise get:
    # Error in names(dat) <- object$term : 'names' attribute [2] must be the same length as the vector [0]
cordat$Plot_Name <- as.factor(cordat$Plot_Name)
cordat$coreID <- as.factor(cordat$coreID)
cordat$species <- as.factor(cordat$species)
cordat$crown_fac <- as.factor(ifelse(cordat$Crown_Class == 3, "Codom", "Inter"))
cordat$year_fac <- as.factor(cordat$Year)

num_check <- cordat %>% select(Plot_Name, coreID, species, Crown_Class) %>% unique() %>% 
  group_by(species, Crown_Class) %>% tally()

num_check


```

Initially I was going to run a model that included all species, but I had to drop crown class, and the model would have been too complex assuming species respond differently over time. Instead, going with species-specific models by crown class.

```{r gam_picrub}
table(cordat$species, cordat$Crown_Class) 
# can only use 1 core per plot and species for hierarchical modelling, so dropping any cores younger than 1984 (acid dep) 
# then randomly selecting one with sample_n

piru84 <- cordat %>% filter(cordat$species == "PIRU") %>% filter(Year == 1984) %>% select(coreID) %>% unique()

piru_cores <- cordat %>% filter(coreID %in% piru84$coreID) %>% droplevels() %>% group_by(Plot_Name) %>% 
  sample_n(1) %>% select(coreID) %>% ungroup()

piru <- cordat %>% filter(coreID %in% piru_cores$coreID)

head(piru)

check <- piru %>% select(Plot_Name,crown_fac, coreID) %>% unique() %>% group_by(Plot_Name, crown_fac) %>% 
  summarize(corecnt = n()) %>% filter(corecnt > 1) %>% nrow()# none

check
e <- piru %>% select(BAIcm2) %>% filter(BAIcm2 != 0) %>% slice(which.min(BAIcm2)) %>% as.numeric()
e #-2.052804
piru <- piru %>% mutate(baishift = BAIcm2 + -1*e + 0.00001,
                        logBAI = log(baishift)) %>% filter(logBAI > -10) # remove 1 large negative outlier

piru <- piru[,c(1:39, 117:118, 112:116, 40:111)]

# Check on how believable linear relationship is for non-smoothed covariates
# ggplot(piru, aes(x = live_stems_ha, y = logBAI))+
#   geom_point()+
#   geom_smooth()
# 
# ggplot(piru, aes(x = elev_m, y = logBAI))+
#   geom_point()+
#   geom_smooth()
# 
# ggplot(piru, aes(x = northiness, y = logBAI))+
#   geom_point()+
#   geom_smooth()

gam_piru1 <- gam(logBAI ~ s(Year, by = crown_fac, k = 50) + 
                           te(northiness, elev_m) + 
                           live_stems_ha +
                         s(Plot_Name, bs = 're', k = 10) + 
                         s(year_fac, bs = 're', k = 100),
                data = piru, method = "REML")

k.check(gam_piru1) #looks ok
appraise(gam_piru1) #not bad? less variance in tails b/c fewer samples
gratia::draw(gam_piru1, residuals = T, rug = T)
draw(gam_piru1)
#str(gam_piru)
summary(gam_piru)
summary(gam_piru1)


gam_piru <- gam(logBAI ~ s(Year, by = crown_fac, k = 50) + 
                           northiness + 
                           live_stems_ha +
                         s(Plot_Name, bs = 're', k = 10) + 
                         s(year_fac, bs = 're', k = 100),
                data = piru, method = "REML")

k.check(gam_piru) #looks ok
appraise(gam_piru) #not bad? less variance in tails b/c fewer samples
gratia::draw(gam_piru, residuals = T, rug = T)
draw(gam_piru)
#str(gam_piru)
summary(gam_piru)
summary(gam_piru1)

gam_piru$aic # 8112 no int
gam_piru1$aic # 8136 int

# piruviz <- getViz(gam_piru)
# piruviz
# str(piruviz)
# p <- plot(sm(piruviz, 1))
     
broom::tidy(gam_piru)
new_dat <- piru %>% select(Plot_Name, logBAI, Year, year_fac, crown_fac, northiness, live_stems_ha)

pred <- piru %>% select(Plot_Name, logBAI, Year, year_fac, crown_fac, northiness, live_stems_ha) %>%                       bind_cols(as_tibble(predict(gam_piru, new_dat, se.fit = T))) %>%    
                 select(Plot_Name, Year, year_fac, crown_fac, northiness, live_stems_ha, fit, se.fit) %>% 
                 mutate(Year = as.numeric(Year)) %>% unique() %>% data.frame()
names(piru)  
str(pred)
head(pred)

p <- ggplot(pred, aes(x = Year, y = fit, fill = crown_fac, color = crown_fac)) + 
       #geom_ribbon(aes(ymin = fit - se.fit, ymax = fit + se.fit), alpha = 0.2) +
       #geom_line() + 
       forestNETN::theme_FHM() +
       geom_smooth()

p

p2 <- ggplot(pred, aes(x = Year, y = fit, group = crown_fac))+
  geom_line(aes(group = Plot_Name), alpha = 0.1) +
  geom_point(alpha = 0.1) + 
  # geom_ribbon(aes(ymin = fit - 1.96*se.fit, 
  #                 ymax = fit + 1.96*se.fit),
  #             alpha = 0.4, fill = 'red')
  
p2

print(plot(piruviz, allTerms = T, seWithMean = 1), pages = 1)
print(plot(piruviz, select = 1:2, ylab = "log(BAI)"), pages = 1)


gam_piru2 <- gam(logBAI ~ s(Year, by = crown_fac, k = 50) + 
                            northiness + 
                            live_stems_ha +
                          te(SPEI03_8, Year) +
                          s(Plot_Name, bs = 're', k = 10),
                 data = piru, method = "REML")


p <- ggplot(piru, aes(x = Year, y = logBAI, group = crown_fac)) 

p + stat_smooth(method = 'gam')


gam_piru$aic #5889.583
gam_piru2$aic # 5853.478 winner

k.check(gam_piru2)
appraise(gam_piru2) #not bad?
gratia::draw(gam_piru2, residuals = T, rug = T)
draw(gam_piru)
str(gam_piru)
summary(gam_piru)

# not very helpful
sm_diffs <- difference_smooths(gam_piru, smooth = 's(Year)')
draw(sm_diffs)

sms <- evaluate_smooth(gam_piru2, "s(Year)") # gives estimate for each year by crown fac
draw(sms)
head(sms)

ggplot(sms, aes(x = Year, y = est, group = crown_fac, color = crown_fac, fill = crown_fac))+
  geom_ribbon(aes(ymin = est - se, ymax = est + se), alpha = 0.2)+
  scale_fill_discrete(name = "Crown Class", "3" = )+
  scale_color_discrete()+
  geom_line() + theme_FHM()
?scale_fill_discrete
draw(gam_piru2)

```

