---
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 6, scipen = 100)
```

<h5> GAMMs to model raw ring width growth of Red spruce (<i>Picea rubens</i>)</h5> {.tabset}
### Compile Data
```{r import}
library(tidyverse)
library(GGally) #for sploms/corr checking
library(mgcv)
library(gratia) # easy plotting, but couldn't get main plotting functions to work 
library(mgcViz) # more customized gam plotting
library(corrr) # correlation matrix
library(forecast) # for auto.arima
library(corrplot) # for inspecting collinearity
library(tseries) # for Dickey-Fuller test of stationarity
library(itsadug) # for gam diagnostics
ctrl_nlme <- nlme::nlmeControl(maxIter = 1000, msmaxEval = 1000, msVerbose = TRUE)
ctrl_gam <- gam.control(trace = TRUE, nthreads = 4, ncv.threads = 4, maxit = 1000,
                        nlm = list(ctrl_nlme)) #// processing

# Used Compiling_covariates.R to compile the main dataset.
cordat_pre <- read.csv("./data/ACAD_final_core_data_20240315.csv") 
table(cordat_pre$Crown_Class)
names(cordat_pre)
cordat <- cordat_pre %>% filter(Crown_Class %in% c(3, 4)) 
table(cordat$Crown_Class)

# have to convert to factors, otherwise get:
    # Error in names(dat) <- object$term : 'names' attribute [2] must be the same length as the vector [0]
cordat$Plot_Name <- as.factor(cordat$Plot_Name)
cordat$coreID <- as.factor(cordat$coreID)
cordat$species <- as.factor(cordat$species)
cordat$crown_fac <- as.factor(ifelse(cordat$Crown_Class == 3, "Codom", "Inter"))
cordat$year_fac <- as.factor(cordat$Year)
str(cordat$crown_fac)
table(cordat$crown_fac, cordat$Crown_Class)

table(cordat$forest_type)

num_check <- cordat %>% select(Plot_Name, coreID, species, Crown_Class) %>% unique() %>% 
  group_by(species, Crown_Class) %>% tally()

num_check
```

# Split data into training/testing datasets for cross-validation
```{r}
set.seed(20240227)
piru_codom1 <- cordat |> filter(species == 'PIRU') |> filter(Crown_Class == 3)

piru_cores <- data.frame(coreID = sort(unique(piru_codom1$coreID)))
samp <- sample.int(n = nrow(piru_cores), size = floor(0.90*nrow(piru_cores)), replace = F)
piru_core_train <- piru_cores[samp,]

piru_codom_full <- piru_codom1 |> 
  mutate(Training = ifelse(coreID %in% piru_core_train, "Train", "Test")) 

table(piru_codom_full$Training)
#415/(3246+415) #~10%

# can only use 1 core per plot and species for hierarchical modelling, so dropping any cores younger than 1984 (acid dep) then randomly selecting one with sample_n
piru84 <- piru_codom_full %>% filter(between(Year, 1984, 2020)) %>% select(coreID) %>% unique()

table(piru_codom_full$Plot_Name)

piru_cores <- piru_codom_full %>% filter(coreID %in% piru84$coreID) %>% 
  select(Plot_Name, coreID) |> group_by(Plot_Name) |> unique() |> 
  sample_n(1) |>  
  ungroup() 

nrow(piru_cores) #73

piru_codom_full$bai_cent = piru_codom_full$BAIcm2 - mean(piru_codom_full$BAIcm2)

e <- piru_codom_full %>% select(BAIcm2) %>% filter(BAIcm2 != 0) %>% 
  slice(which.min(BAIcm2)) %>% as.numeric()
e #0.3

hist(piru_codom_full$BAIcm2)
hist(piru_codom_full$bai_cent)

std01 <- function(x){(x-min(x))/(max(x)-min(x))}
cent <- function(x){x - mean(x)}

piru_codom_full <- piru_codom_full %>% 
  mutate(logBAI = log(BAIcm2 + e), # remove 1 large negative outlier with log
         year_cent = Year - min(Year),
         elev_cent = cent(elev_m),
         elev_std = std01(elev_m),
         dbh_cent = cent(DBH),
         dbh_std = std01(DBH),
         live_stems_cent = cent(live_stems_ha),
         live_stems_std = std01(live_stems_ha),
         est_age_cent = cent(est_age),
         est_age_std = std01(est_age)
)

piru_codom1 <- piru_codom_full %>% filter(coreID %in% piru_cores$coreID) |> 
  filter(!is.na(BAIcm2)) # drop plots missing BAIcm2 and only include 1 core per plot

length(unique(piru_codom1$coreID)) #73

check <- piru_codom1 %>% select(Plot_Name, crown_fac, coreID) %>% 
  unique() %>% group_by(Plot_Name, crown_fac) %>% 
  summarize(corecnt = n()) %>% filter(corecnt > 1) %>% nrow()# none

check # 0 all plots only have 1 core

piru_codom <- piru_codom1 |> filter(Training == "Train") |> droplevels()
piru_codom_test1 <- piru_codom1 |> filter(Training == "Test") |> droplevels()
piru_codom_test2 <- piru_codom_full |> filter(!coreID %in% piru_cores$coreID) |> droplevels()

table(piru_codom$coreID)
table(piru_codom_test1$coreID)
table(piru_codom_test2$coreID)
```

### Exploratory Data Analysis {.tabset}
```{r}
cormat_bai <- piru_codom |> select_if(is.numeric) |> correlate() |> 
   focus(BAIcm2) |> filter(abs(BAIcm2) >= 0.15) 

ggplot(piru_codom, aes(x = Year, y = BAIcm2, color = coreID)) + 
  geom_point() + geom_smooth(se = F) + theme(legend.position = 'none') + 
  forestNETN::theme_FHM() + scale_color_viridis_d()

# ggpairs(piru_codom |> select(BAIcm2,
#                              BA_pct_lg, est_age, live_stems_ha, DBH, northiness,
#                              pctTC, C_N, horizon_depth, Ca_Al, soilpH,
#                              NO3_roll5, pH_roll5,
#                            	 spring_5c_days_roll5, 
#                              SPEI01_5_roll4, SPEI03_10_roll5, 
#                              #ppt_4_roll5, ppt_8_roll5, 
#                              ppt_gs_roll5,
#                              ))

cor_mat <- 
round(cor(piru_codom[,c("BAIcm2", "BA_pct_lg", "est_age", "live_stems_ha", "DBH", 'northiness', 
                        'elev_m', 'pctTC', 'C_N', 'horizon_depth', 'Ca_Al', 'soilpH',
                        'NO3_roll5', 'pH_roll5', "SO4_roll5",
                        'spring_5c_days_roll5', 'SPEI01_5_roll4', 'SPEI03_10_roll5', 'SPEI03_8_roll5',
                        'ppt_gs_roll5', 'ppt_8_roll5')], 
          use = 'pairwise.complete.obs'),2)

corrplot(cor_mat, method = 'circle')

# Checking that climate vars make sense
ggplot(piru_codom, aes(x = Year, y = SPEI01_5)) +
  geom_point() + geom_smooth(se = F) + 
  forestNETN::theme_FHM() + 
  geom_point(aes(y = SPEI01_5_roll5), color = 'grey') +
  geom_smooth(aes(y = SPEI01_5_roll5, x = Year), color = 'dimgrey', se = F) +
  scale_x_continuous(breaks = seq(1980, 2020, by = 2))

ggplot(piru_codom, aes(x = Year, y = ppt_4_roll5)) +
  geom_point() + geom_smooth() + forestNETN::theme_FHM()

ggplot(piru_codom, aes(x = Year, y = tmax_gs_roll5)) +
  geom_point() + geom_smooth() + forestNETN::theme_FHM()

ggplot(piru_codom, aes(x = Year, y = gs_5c_length_roll5)) +
  geom_point() + geom_smooth() + forestNETN::theme_FHM()

ggplot(piru_codom, aes(x = Year, y = tmin_wint_roll5)) +
  geom_point() + geom_smooth() + forestNETN::theme_FHM()

ggplot(piru_codom, aes(x = Year, y = NO3_roll3)) +
  geom_point() + geom_smooth() + forestNETN::theme_FHM()

ggplot(piru_codom, aes(x = Year, y = BAIcm2, color = forest_type, fill = forest_type)) + 
  geom_point(alpha = 0.2) + geom_smooth() +
  theme(legend.position = 'right') + forestNETN::theme_FHM()

ggplot(piru_codom, aes(x = Year, y = BAIcm2, color = factor(fire1947), 
                       fill = factor(fire1947))) + 
  geom_point(alpha = 0.2) + geom_smooth() +
  theme(legend.position = 'right') + forestNETN::theme_FHM()

```


```{r piru_gam}

#write.csv(piru_codom, "./data/PIRU_codom_only_dataset_BAI.csv")

hist(piru_codom$elev_m)
hist(piru_codom$DBH)
hist(piru_codom$rrw_cent)
hist(piru_codom$northiness)
hist(piru_codom$PlotSlope)

ggplot(piru_codom, aes(x = elev_std, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = soilpH, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = BA_pct_lg, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = DBH, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = est_age, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = PlotSlope, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = northiness, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = pctTC, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = horizon_depth, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = year_cent, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = ppt_8_roll5, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = gs_5c_length_roll5, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = NO3_roll5, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = SO4_roll5, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = ppt_gs_roll5, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = SPEI01_5_roll5, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = pH_roll5, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = SPEI03_10_roll5, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = SPEI03_10_roll5, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = gs_5c_length_roll5, y = logBAI)) + geom_point() + geom_smooth() + forestNETN::theme_FHM()

# ti b/c only including interaction terms and including year_cent as main effect separately
gam_piru <- gam(logBAI ~ forest_type + # dropped gamm b/c wouldn't converge
                       #BA_pct_lg + #northiness +
                          dbh_std + # correlated with BA_pct_lg
                          horizon_depth + elev_std + 
                          soilpH + Ca_Al +
                          s(year_cent, k = 5) + 
                          ti(gs_5c_length_roll5, year_cent, k = c(10, 10)) +
                          #ti(SPEI01_5_roll5, year_cent, k = c(10, 10)) +
                          ti(ppt_8_roll5, year_cent, k = c(5, 5)) + # less correlated with SO4 than gs
                          ti(SO4_roll5, year_cent, k = c(5, 5)) +
                          s(Plot_Name, bs = 're', k = 40) +   
                          s(year_fac, bs = 're', k = 5),
                 family = gaussian, 
                # family = tw(), #tweedie dist, will determine p 
                data = piru_codom, method = "REML", control = ctrl_gam)

gam.check(gam_piru)
summary(gam_piru)
par(mfrow = c(2,3))
plot(gam_piru$gam)
par(mfrow = c(1,1))

acf(resid(gam_piru)) # acf drops below 0 at 6; q = 6 
pacf(resid(gam_piru)) #pcf drops below 0 at 2; p = 2
adf.test(piru_codom$logBAI) #lagorder = 13 Dickey-Fuller = -6.8907
arma_res <- auto.arima(resid(gam_piru), 
                       stationary = FALSE, # b/c adf.test rejects stationarity
                       seasonal = FALSE, 
                       stepwise = FALSE, approximation = FALSE) # suggested by help info
arma_res # order is p = order of ar model, d = order of differencing, q = order of moving average
# returns ar1, ar2, meaning p = 2 

gam_piru_arma <- gam(logBAI ~ forest_type + # dropped gamm b/c wouldn't converge
                          dbh_std + # correlated with BA_pct_lg
                          horizon_depth + elev_std + 
                          soilpH + Ca_Al +
                          s(year_cent, k = 5) + 
                          ti(gs_5c_length_roll5, year_cent, k = c(15, 15)) +
                          ti(ppt_8_roll5, year_cent, k = c(5, 5)) + # less correlated with SO4 than gs
                          ti(SO4_roll5, year_cent, k = c(5, 5)) +
                          s(Plot_Name, bs = 're', k = 40) +   
                          s(year_fac, bs = 're', k = 5),
                 family = gaussian, 
                 correlation = corARMA(form = ~1|year_fac, p = 2), 
                 data = piru_codom, method = "REML", control = ctrl_gam)

anova(gam_piru, gam_piru_arma)#

par(mfrow = c(2,2))
gam.check(gam_piru_arma)
par(mfrow = c(1,1))
summary(gam_piru_arma)

acf_resid(gam_piru_arma)

plot(gam_piru_arma, rug = T, residuals = T, select = 1, pch = 1, cex = 1)
plot(gam_piru_arma, rug = T, residuals = T, select = 1, cex = 0.5, pch = 1,
     shade = T, seWithMean = T, shift = coef(gam_piru_arma)[1])

#gratia::appraise(gam_piru_arma$gam) # Error- not sure why
#gratia::draw(gam_piru) # Error- not sure why

g <- getViz(gam_piru_arma)
p <- plot(sm(g, 1))
check(g)

p + 
  l_fitLine(color = 'dodgerblue2', lwd = 0.5) + 
  l_rug() +
  l_ciLine(color = "dodgerblue3", lwd = 1) +
  l_points(shape = 19, size = 1, alpha = 0.2)

listLayers(p)

plot(sm(g, 2)) + l_fitRaster() + l_fitContour() + l_points() + 
  l_rug() +
  forestNETN::theme_FHM()
plot(sm(g, 3)) + l_fitRaster() + l_fitContour() + l_points() + 
  l_rug() +
  forestNETN::theme_FHM()
plot(sm(g, 4)) + l_fitRaster() + l_fitContour() + l_points() + 
  l_rug() +
  forestNETN::theme_FHM()

summary(g)

plot(g, select = 1) + l_dens(type = "cond") + l_fitLine() + l_ciLine()

qq(g, rep = 20, showReps = T, CI = "none", 
   a.qqpoi = list("shape" = 19), a.replin = list("alpha" = 0.2))

check(g,
      a.qq = list(method = "tnorm", 
                  a.cipoly = list(fill = "light blue")), 
      a.respoi = list(size = 0.5), 
      a.hist = list(bins = 10))

ggplot(piru_codom, aes(x = Year, y = SO4_roll5)) + 
  geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = Year, y = ppt_8_roll5)) + 
  geom_point() + geom_smooth() + forestNETN::theme_FHM()
ggplot(piru_codom, aes(x = Year, y = gs_5c_length_roll5)) + 
  geom_point() + geom_smooth() + forestNETN::theme_FHM()

summary(gam_piru_arma)
vis.gam(gam_piru_arma, view = c("year_cent", "ppt_8_roll5"), plot.type = "contour", color = 'topo')

pacf(resid(gam_piru$lme, type = 'normalized'))
pacf(resid(gam_piru_arma$lme, type = 'normalized'))
# I don't understand why the arma pacf looks worse

acf(resid_gam(gam_piru_arma))
plot_modelfit(gam_piru_arma, view = "year_cent")
plot_modelfit(gam_piru_arma, view = "SO4_roll5")

par(mfrow = c(2,3))
diagnostics(gam_piru_arma, plot = 1:3)
par(mfrow = c(1,1))

checkresiduals(gam_piru_arma, plot = T, test = "LB")
par(mfrow = c(1,1))

vis.gam(gam_piru_arma, type = 'response', plot.type = 'parsp')

```
### Cross validation

```{r}
CVgam
```

Next steps
1. Visualize results better
2. More diagnostics (check tweedie without log)
3. Cross validate with cores dropped from plots
4. Cross validate with plots dropped from analysis
Next I need to generate rasters of ACAD with the values of the model and use model to predict.