---
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 6, scipen = 100)
```

<h5> GAMMs to model raw ring width growth of Red spruce (<i>Picea rubens</i>)</h5> {.tabset}
### Compile Data
```{r import}
library(tidyverse)
library(GGally) #for sploms/corr checking
library(mgcv)
library(gratia) # easy plotting, but couldn't get main plotting functions to work 
library(mgcViz) # more customized gam plotting
library(corrr) # correlation matrix

ctrl <- gam.control(trace = TRUE) # So see ~ progress of gam

# Used Compiling_covariates.R to compile the main dataset.
cordat_pre <- read.csv("./data/ACAD_final_core_data_20240311.csv") 
table(cordat_pre$Crown_Class)
names(cordat_pre)
cordat <- cordat_pre %>% filter(Crown_Class %in% c(3, 4)) 
table(cordat$Crown_Class)

# have to convert to factors, otherwise get:
    # Error in names(dat) <- object$term : 'names' attribute [2] must be the same length as the vector [0]
cordat$Plot_Name <- as.factor(cordat$Plot_Name)
cordat$coreID <- as.factor(cordat$coreID)
cordat$species <- as.factor(cordat$species)
cordat$crown_fac <- as.factor(ifelse(cordat$Crown_Class == 3, "Codom", "Inter"))
cordat$year_fac <- as.factor(cordat$Year)
str(cordat$crown_fac)
table(cordat$crown_fac, cordat$Crown_Class)

table(cordat$forest_type)

num_check <- cordat %>% select(Plot_Name, coreID, species, Crown_Class) %>% unique() %>% 
  group_by(species, Crown_Class) %>% tally()

num_check
```

# Split data into training/testing datasets for cross-validation
```{r}
set.seed(20240227)
piru_codom1 <- cordat |> filter(species == 'PIRU') |> filter(Crown_Class == 3)
piru_cores <- data.frame(coreID = sort(unique(piru_codom1$coreID)))
samp <- sample.int(n = nrow(piru_cores), size = floor(0.80*nrow(piru_cores)), replace = F)
piru_core_train <- piru_cores[samp,]

piru_codom_full <- piru_codom1 |> 
  mutate(Training = ifelse(coreID %in% piru_core_train, "Train", "Test")) 
table(piru_codom_full$Training)
#845/(3246+845) #~20%

piru_codom <- piru_codom_full |> filter(Training == "Train")

```

### Exploratory Data Analysis {.tabset}
```{r}
cormat_rrw <- piru_codom |> select_if(is.numeric) |> correlate() |> 
  focus(RRWmm) |> filter(abs(RRWmm) >= 0.07)

cormat_bai <- piru_codom |> select_if(is.numeric) |> correlate() |> 
   focus(BAIcm2) |> filter(abs(BAIcm2) >= 0.1)

ggplot(piru_codom, aes(x = Year, y = RRWmm)) + geom_point() + geom_smooth() + 
  forestNETN::theme_FHM()


ggpairs(piru_codom |> select(RRWmm,
                             est_age, pctTC, horizon_depth, C_N, Pct_Bryophyte,
                             liveBA_m2ha, BA_pct_lg, pctTC, Ca_Al, DBH, 
                             ppt_8_roll5, ppt_gs_roll5, SPEI01_8_roll5, SPEI03_8_roll5,
                             soilpH, Pct_Rock, elev_m))

cormat <- piru_codom |> select(RRWmm,
                               est_age, pctTC, horizon_depth, C_N, Pct_Bryophyte,
                               liveBA_m2ha, BA_pct_lg, pctTC, Ca_Al, DBH, 
                               ppt_8_roll5, ppt_gs_roll5, SPEI01_8_roll5,
                               SPEI03_8_roll5, 
                               gs_ff_length_roll5,
                               gs_5c_length_roll5,
                               spring_5c_days_roll5, 
                               spring_5c_days, 
                               spring_5c_days_lag1,
                               fall_5c_days_lag1,
                               fall_5c_days_roll5,
                               soilpH, Pct_Rock, elev_m, pH, pH_roll5) |> correlate()

ggplot(piru_codom, aes(x = Year, y = SPEI03_9_roll3)) +
  geom_point() + geom_smooth() + forestNETN::theme_FHM() + 
  scale_x_continuous(breaks = seq(1980, 2020, by = 2))

ggplot(piru_codom, aes(x = Year, y = SPEI03_9)) +
  geom_point() + geom_smooth() + forestNETN::theme_FHM()

ggplot(piru_codom, aes(x = Year, y = tmax_gs_roll3)) +
  geom_point() + geom_smooth() + forestNETN::theme_FHM()

ggplot(piru_codom, aes(x = Year, y = gs_5c_length_roll5)) +
  geom_point() + geom_smooth() + forestNETN::theme_FHM()

ggplot(piru_codom, aes(x = Year, y = tmin_wint_roll3)) +
  geom_point() + geom_smooth() + forestNETN::theme_FHM()

ggplot(piru_codom, aes(x = Year, y = NO3_roll3)) +
  geom_point() + geom_smooth() + forestNETN::theme_FHM()

ggplot(piru_codom, aes(x = Year, y = RRWmm, color = forest_type, fill = forest_type)) + 
  geom_point(alpha = 0.2) + geom_smooth() +
  theme(legend.position = 'right') + forestNETN::theme_FHM()

ggplot(piru_codom, aes(x = Year, y = RRWmm, color = factor(fire1947), 
                       fill = factor(fire1947))) + 
  geom_point(alpha = 0.2) + geom_smooth() +
  theme(legend.position = 'right') + forestNETN::theme_FHM()

ggplot(piru_codom, aes(x = Year, y = BAIcm2, color = forest_type, fill = forest_type)) +   geom_point(alpha = 0.2) + geom_smooth() +
  theme(legend.position = 'right') + forestNETN::theme_FHM()

```


```{r piru_gam}
# can only use 1 core per plot and species for hierarchical modelling, so dropping any cores younger than 1984 (acid dep) then randomly selecting one with sample_n
table(piru_codom$Year)
hist(piru_codom$Year) # most cores cover 1984 to 2018

piru84 <- piru_codom %>% filter(between(Year, 1984, 2020)) %>% select(coreID) %>% unique()
head(piru84)

piru_cores <- piru_codom %>% filter(coreID %in% piru84$coreID) %>% 
  droplevels() %>%  group_by(Plot_Name) %>% 
  sample_n(1) %>% select(Plot_Name, coreID) %>% ungroup() 

piru_codom <- piru_codom %>% filter(coreID %in% piru_cores$coreID) |> 
  filter(!is.na(RRWmm)) # drop plot 41 missing RRW

check <- piru_codom %>% select(Plot_Name,crown_fac, coreID) %>% 
  unique() %>% group_by(Plot_Name, crown_fac) %>% 
  summarize(corecnt = n()) %>% filter(corecnt > 1) %>% nrow()# none

check # all plots only have 1 core

piru_codom$rrw_cent = piru_codom$RRWmm - mean(piru_codom$RRWmm)

e <- piru_codom %>% select(RRWmm) %>% filter(RRWmm != 0) %>% 
  slice(which.min(RRWmm)) %>% as.numeric()
e #0.088

hist(piru_codom$RRWmm)
hist(piru_codom$rrw_cent)

std01 <- function(x){(x-min(x))/(max(x)-min(x))}
cent <- function(x){x - mean(x)}

piru_codom <- piru_codom %>% 
  mutate(logRRW = log(RRWmm + e), # remove 1 large negative outlier with log
         year_cent = Year - min(Year),
         elev_cent = cent(elev_m),
         elev_std = std01(elev_m),
         dbh_cent = cent(DBH),
         dbh_std = std01(DBH),
         live_stems_cent = cent(live_stems_ha),
         live_stems_std = std01(live_stems_ha),
         est_age_cent = cent(est_age),
         est_age_std = std01(est_age)
)

write.csv(piru_codom, "./data/PIRU_codom_only_dataset.csv")

ggplot(piru_codom, aes(x = Year, y = RRWmm, group = coreID, color = forest_type)) +
  geom_line() + forestNETN::theme_FHM() + scale_color_viridis_d()

# Trying another model with more soils
# Using first model to determine optimal ARMA
hist(piru_codom$elev_m)
hist(piru_codom$DBH)
hist(piru_codom$rrw_cent)
hist(piru_codom$northiness)
hist(piru_codom$PlotSlope)

ggplot(piru_codom, aes(x = elev_std, y = logRRW)) + geom_point() + geom_smooth()
ggplot(piru_codom, aes(x = year_cent, y = logRRW)) + geom_point() + geom_smooth()
ggplot(piru_codom, aes(x = pctTC, y = logRRW)) + geom_point() + geom_smooth()
ggplot(piru_codom, aes(x = horizon_depth, y = logRRW)) + geom_point() + geom_smooth()
ggplot(piru_codom, aes(x = ppt_8_roll5, y = logRRW)) + geom_point() + geom_smooth()
ggplot(piru_codom, aes(x = gs_5c_length_roll5, y = logRRW)) + geom_point() + geom_smooth()

gam_piru <- gamm(logRRW ~ forest_type + #BA_pct_lg +  
                          fire1947 + elev_std +
                          dbh_std + 
                          #est_age_std + 
                          pctTC + #horizon_depth + soilpH +
                          s(year_cent, k = 10) + 
                          te(SPEI03_8_roll5, year_cent, k = c(10,10)) +
                          te(gs_5c_length_roll5, year_cent, k = c(10, 10)) +
                          te(NO3_roll5, year_cent, k = c(10, 10)) +
                          s(Plot_Name, bs = 're', k = 40) +   
                             # each plot/crown combination has its own intercept
                          s(year_fac, bs = 're', k = 5),
                 family = gaussian, 
                # family = tw(), #tweedie dist, will determine p 
                data = piru_codom, method = "REML")#, control = ctrl)

gam_piru_ar6 <- gamm(logRRW ~ forest_type + #BA_pct_lg +  
                          fire1947 + elev_std +
                          dbh_std + 
                          #est_age_std + 
                          pctTC + #horizon_depth + soilpH +
                          s(year_cent, k = 10) + 
                          te(SPEI03_8_roll5, year_cent, k = c(10,10)) +
                          te(gs_5c_length_roll5, year_cent, k = c(10, 10)) +
                          te(NO3_roll5, year_cent, k = c(10, 10)) +
                          s(Plot_Name, bs = 're', k = 40) +   
                             # each plot/crown combination has its own intercept
                          s(year_fac, bs = 're', k = 5),
                 family = gaussian, 
                # family = tw(), #tweedie dist, will determine p 
                correlation = corARMA(form = ~1|Year, p = 6),
                data = piru_codom, method = "REML")#, control = ctrl)

# str(gam_piru_ar0s)
gam.check(gam_piru$gam)
summary(gam_piru$gam)
par(mfrow = c(2,3))
plot(gam_piru$gam)
par(mfrow = c(1,1))

acf(resid(gam_piru$lme))
pacf(resid(gam_piru$lme))
acf(resid(gam_piru_ar6$lme))
pacf(resid(gam_piru_ar6$lme))

#plot(gam_piru$gam)

# gratia::appraise(gam_piru) # Error- not sure why
# gratia::draw(gam_piru) # Error- not sure why

plot(gam_piru$gam, scale = 0)

g <- getViz(gam_piru)
#plot(sm(g, 6))
#print(plot(g, allTerms = T))
check(g)

vis.gam(gam_piru, view = c("year_cent", "ppt_8_roll5"), plot.type = "contour", color = 'topo')
vis.gam(gam_piru, view = c("ppt_8_roll5", "elev_std"), plot.type = 'contour')
vis.gam(gam_piru, view = c('ppt_8_roll5', "dbh_std"), plot.type = 'contour')

#plot(g, residuals = TRUE)

b0 <- coef(gam_piru)[1]
draw(gam_piru, residuals = T)

plot.gam(gam_piru)

```

Next steps
1. Visualize results better
2. More diagnostics (check tweedie without log)
3. Cross validate with cores dropped from plots
4. Cross validate with plots dropped from analysis
Next I need to generate rasters of ACAD with the values of the model and use model to predict.