---
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 6, scipen = 100)
```

<h5> GAMMs to model raw ring width growth of Red spruce (<i>Picea rubens</i>)</h5> {.tabset}
### Compile Data
```{r import}
library(tidyverse)
library(corrr) # correlation matrix
library(corrplot) # for inspecting collinearity
library(nlme)
library(MuMIn)

ctrl_nlme <- nlme::nlmeControl(maxIter = 10000, msMaxIter = 10000, msVerbose = FALSE)

# Used Compiling_covariates.R to compile the main dataset.
cordat_pre <- read.csv("./data/ACAD_final_core_data_20240315.csv") 
cordat <- cordat_pre %>% filter(Crown_Class %in% c(3, 4)) |> filter(!is.na(BAIcm2))
table(cordat$Crown_Class)

num_check <- cordat_pre %>% select(Plot_Name, coreID, species, Crown_Class) %>% unique() %>% 
  group_by(species, Crown_Class) %>% tally()

num_check
```

## Split data into training/testing datasets for cross-validation
```{r}
set.seed(20240227)
piru_codom1 <- cordat |> filter(species == 'PIRU') |> filter(Crown_Class == 3)

piru_cores <- data.frame(coreID = sort(unique(piru_codom1$coreID)))
samp <- sample.int(n = nrow(piru_cores), size = floor(0.90*nrow(piru_cores)), replace = F)
piru_core_train <- piru_cores[samp,]

piru_codom_full <- piru_codom1 |> 
  mutate(Training = ifelse(coreID %in% piru_core_train, "Train", "Test")) 

table(piru_codom_full$Training)
#434/(3657+434) #~10%

# can only use 1 core per plot and species for hierarchical modelling, so dropping any cores younger than 1984 (acid dep) then randomly selecting one with sample_n
piru84 <- piru_codom_full %>% filter(between(Year, 1984, 2020)) %>% select(coreID) %>% unique()

table(piru_codom_full$Plot_Name)

piru_cores <- piru_codom_full %>% filter(coreID %in% piru84$coreID) %>% 
  select(Plot_Name, coreID) |> group_by(Plot_Name) |> unique() |> 
  sample_n(1) |>  
  ungroup() 

nrow(piru_cores) #73

piru_codom_full$bai_cent = piru_codom_full$BAIcm2 - mean(piru_codom_full$BAIcm2)

e <- piru_codom_full %>% select(BAIcm2) %>% filter(BAIcm2 != 0) %>% 
  slice(which.min(BAIcm2)) %>% as.numeric()
e #0.3

#hist(piru_codom_full$BAIcm2)
#hist(piru_codom_full$bai_cent)

std01 <- function(x){(x-min(x))/(max(x)-min(x))}
cent <- function(x){x - mean(x)}

piru_codom_full <- piru_codom_full %>% 
  mutate(logBAI = log(BAIcm2 + e), # remove 1 large negative outlier with log
         year_cent = Year - min(Year),
         elev_cent = cent(elev_m),
         elev_std = std01(elev_m),
         dbh_cent = cent(DBH),
         dbh_std = std01(DBH),
         live_stems_cent = cent(live_stems_ha),
         live_stems_std = std01(live_stems_ha),
         est_age_cent = cent(est_age),
         est_age_std = std01(est_age)
)

# have to convert to factors, otherwise get:
    # Error in names(dat) <- object$term : 'names' attribute [2] must be the same length as the vector [0]
piru_codom_full$Plot_Name <- as.factor(piru_codom_full$Plot_Name)
piru_codom_full$coreID <- as.factor(piru_codom_full$coreID)
piru_codom_full$species <- as.factor(piru_codom_full$species)
piru_codom_full$year_fac <- as.factor(piru_codom_full$Year)

piru_codom1 <- piru_codom_full %>% filter(coreID %in% piru_cores$coreID) |> 
  filter(!is.na(BAIcm2)) # drop plots missing BAIcm2 and only include 1 core per plot

length(unique(piru_codom1$coreID)) #73

check <- piru_codom1 %>% select(Plot_Name, Crown_Class, coreID) %>% 
  unique() %>% group_by(Plot_Name, Crown_Class) %>% 
  summarize(corecnt = n()) %>% filter(corecnt > 1) %>% nrow()# none

check # 0 all plots only have 1 core

# training dataset; complete 5-year rolling averages start at 1986 for deposition data
piru_codom <- piru_codom1 |> filter(Training == "Train") |> droplevels() |> filter(Year >= 1986)
# first test dataset
piru_codom_test1 <- piru_codom1 |> filter(Training == "Test") |> droplevels() |> filter(Year >= 1986)
# second test dataset
piru_codom_test2 <- piru_codom_full |> filter(!coreID %in% piru_cores$coreID) |> droplevels() |> filter(Year >= 1986)

table(piru_codom$coreID)
table(piru_codom_test1$coreID)
table(piru_codom_test2$coreID)

table(piru_codom$forest_type)
write.csv(piru_codom, "./data/piru_codom_train.csv")

```

### Checking the Data {.tabset}
#### Correlations {.tabset}
```{r}
cormat_bai <- piru_codom |> select_if(is.numeric) |> correlate() |> 
   focus(BAIcm2) |> filter(abs(BAIcm2) >= 0.15) 

cormat_bai
# cor_mat_dep <- round(cor(piru_codom[,
#                             c("BAIcm2", 
#                               "pH_lag1", "pH_roll2", "pH_roll3", "pH_roll4", "pH_roll5",   
#                               "SO4", "SO4_lag1", "SO4_roll2", "SO4_roll3", "SO4_roll4", "SO4_roll5",
#                               "NO3", "NO3_lag1", "NO3_roll2", "NO3_roll3", "NO3_roll4", "NO3_roll5")], 
#                          use = 'pairwise.complete.obs'), 2)
# 
# corrplot(cor_mat_dep, method = 'circle')
# 
# cor_mat_cc <- round(cor(piru_codom[,c("BAIcm2",
#                               "gs_5c_length", "gs_5c_length_lag1", "gs_5c_length_roll2", "gs_5c_length_roll3",
#                                 "gs_5c_length_roll4", "gs_5c_length_roll5",
#                               "spring_5c_days", "spring_5c_days_lag1", "spring_5c_days_roll2", "spring_5c_days_roll3", 
#                               "spring_5c_days_roll4", "spring_5c_days_roll5",
#                               "tmin_wint", "tmin_wint_lag1", "tmin_wint_roll2", "tmin_wint_roll3", 
#                               "tmin_wint_roll4", "tmin_wint_roll5",
#                               "ppt_gs", "ppt_gs_lag1", "ppt_gs_roll2", "ppt_gs_roll3", "ppt_gs_roll4", "ppt_gs_roll5",
#                               "ppt_8", "ppt_8_lag1", "ppt_8_roll2", "ppt_8_roll3", "ppt_8_roll4", "ppt_8_roll5")],
#                            use = 'pairwise.complete.obs'), 2)
# 
# corrplot(cor_mat_cc, method = 'circle')
# 
# 
# names(piru_codom)
cor_mat <-
round(cor(piru_codom[,c("BAIcm2", "BA_pct_lg", "est_age", "live_stems_ha", "DBH", 'northiness',
                        'elev_m', 'pctTC', 'C_N', 'horizon_depth', 'Ca_Al', 'soilpH',
                        'NO3_roll5', 'pH_roll5', "SO4_roll5", 'spring_5c_days_roll5',
                        'SPEI01_5_roll4', 'SPEI03_10_roll5', 'SPEI03_8_roll5',
                        'ppt_gs_roll5', 'ppt_8_roll5')],
          use = 'pairwise.complete.obs'),2)

corrplot(cor_mat, method = 'circle')
```

#### EDA
```{r}

ggplot(piru_codom, aes(y = gs_5c_length_roll5, x = year_fac, fill = year_fac)) + geom_violin() + 
  forestNETN::theme_FHM() + theme(legend.position = 'none')
ggplot(piru_codom, aes(y = ppt_8_roll5, x = year_fac, fill = year_fac)) + geom_violin() + 
  forestNETN::theme_FHM() + theme(legend.position = 'none')
ggplot(piru_codom, aes(y = SO4_roll5, x = year_fac, fill = year_fac)) + geom_violin() + 
  forestNETN::theme_FHM() + theme(legend.position = 'none')
ggplot(piru_codom, aes(y = NO3_roll5, x = year_fac, fill = year_fac)) + geom_violin() + 
  forestNETN::theme_FHM() + theme(legend.position = 'none')

# Looking at shape of responses
ggplot(piru_codom, aes(x = Year, y = BAIcm2, color = forest_type, fill = forest_type)) + 
  geom_point(alpha = 0.2) + geom_smooth() + scale_colour_brewer(palette = "Set1") +
  theme(legend.position = 'right') + forestNETN::theme_FHM()

ggplot(piru_codom, aes(x = SO4_roll5, y = BAIcm2, color = Plot_Name, fill = Plot_Name)) + 
  geom_point(alpha = 0.2) + geom_smooth(se = F) + 
  theme(legend.position = 'right') + forestNETN::theme_FHM() + theme(legend.position = 'none')

ggplot(piru_codom, aes(x = ppt_8_roll5, y = BAIcm2, color = Plot_Name, fill = Plot_Name)) + 
  geom_point(alpha = 0.2) + geom_smooth(se=F) +
  theme(legend.position = 'right') + forestNETN::theme_FHM() + theme(legend.position = 'none')

ggplot(piru_codom, aes(x = gs_5c_length_roll5, y = BAIcm2, color = Plot_Name, fill = Plot_Name)) + 
  geom_point(alpha = 0.2) + geom_smooth(se=F) +
  theme(legend.position = 'right') + forestNETN::theme_FHM() + theme(legend.position = 'none')

```

### BAI GAMMS
```{r }
# ti b/c only including interaction terms and including year_cent as main effect separately
lme_piru <- lme(logBAI ~ #forest_type + 
                          #BA_pct_lg + 
                          dbh_std + # correlated with BA_pct_lg
                          horizon_depth + 
                          elev_std + 
                          soilpH + #Ca_Al +
                          gs_5c_length_roll5 +
                          poly(ppt_8_roll5, 3) + 
                          SO4_roll5,
                #correlation = corAR1(form = ~1|year_std),
                random = ~1|Plot_Name,
                # family = tw(), #tweedie dist, will determine p 
                data = piru_codom, method = "REML")

summary(lme_piru)
plot(lme_piru, resid(., scaled = TRUE) ~ fitted(.), abline = 0,
     pch = 20, xlab = "Fitted values", ylab = "Standardised residuals")

re <- ranef(lme_piru)[,1]
qqnorm(re)


qqnorm(resid(lme_piru))

mod_df <- data.frame(lme_piru$data |> select(forest_type, dbh_std, horizon_depth, elev_std,
                                             soilpH, gs_5c_length_roll5, ppt_8_roll5, 
                                             SO4_roll5), resid = lme_piru$residuals)

ggplot(mod_df, aes(x = dbh_std, y = resid.fixed)) + geom_point() + forestNETN::theme_FHM()
ggplot(mod_df, aes(x = horizon_depth, y = resid.fixed)) + geom_point() + forestNETN::theme_FHM()
ggplot(mod_df, aes(x = elev_std, y = resid.fixed)) + geom_point() + forestNETN::theme_FHM()
ggplot(mod_df, aes(x = soilpH, y = resid.fixed)) + geom_point() + forestNETN::theme_FHM()
ggplot(mod_df, aes(x = gs_5c_length_roll5, y = resid.fixed)) + geom_point() + forestNETN::theme_FHM()
ggplot(mod_df, aes(x = ppt_8_roll5, y = resid.fixed)) + geom_point() + forestNETN::theme_FHM()
ggplot(mod_df, aes(x = SO4_roll5, y = resid.fixed)) + geom_point() + forestNETN::theme_FHM()

acf(resid(lme_piru, type = 'normalized'))
summary(lme_piru)


```

### Visualizing GAM results
```{r}
#gratia::appraise(gam_piru_arma$gam) # Error- not sure why
#gratia::draw(gam_piru) # Error- not sure why

g <- getViz(gam_piru_arma$gam)
check(g)

plot(sm(g, 1)) + 
  l_fitLine(color = 'dodgerblue2', lwd = 0.5) + 
  l_rug() +
  l_ciLine(color = "dodgerblue3", lwd = 1) +
  l_points(shape = 19, size = 1, alpha = 0.2)

plot(sm(g, 2)) + 
  l_fitLine(color = 'dodgerblue2', lwd = 0.5) + 
  l_ciLine(color = "dodgerblue3", lwd = 1) +
  l_points(shape = 19, size = 1, alpha = 0.2)

plot(sm(g, 3)) + 
  l_fitLine(color = 'dodgerblue2', lwd = 0.5) + 
  l_ciLine(color = "dodgerblue3", lwd = 1) +
  l_points(shape = 19, size = 1, alpha = 0.2)

summary(g)

check(g,
      a.qq = list(method = "tnorm", 
                  a.cipoly = list(fill = "light blue")), 
      a.respoi = list(size = 0.5), 
      a.hist = list(bins = 10))

summary(gam_piru_arma$gam)

acf(resid_gam(gam_piru_arma))
#plot_modelfit(gam_piru_arma, view = "SO4_roll5")
```
### Cross validation
```{r}
head(piru_codom_test1)

pred1 <- predict(gam_piru_arma$gam, newdata = piru_codom_test1, type = 'terms')
pred2 <- predict(gam_piru_arma$gam, newdata = piru_codom_test2, type = 'terms')

```


### Next steps
1. Visualize results better
2. More diagnostics (check tweedie without log)
3. Cross validate with cores dropped from plots
4. Cross validate with plots dropped from analysis
5. Generate rasters of covariates in model ACAD to predict various climate scenarios.

